A tool for designing digital filters for the Hankel and Fourier transforms in potential, diffusive, and wavefield modeling

### 2018-06-18 ManuscriptGeophysics2018.tex
In revision; move from shelve to plain txt files

In the following sections we give a brief overview of the methodology, the
theory, and the code, and then show examples of its usage, as well as
presenting new filters for both CSEM and GPR data. The current article falls
therefore into all three mentioned categories by the new application of DLF to
wave phenomena, \new{which means that the DLF can be used to model seismic
wavefields as well,} and by providing new filters with the accompanying code.
The algorithm and many more examples of its usage can be found on
\href{https://github.com/empymod/article-fdesign}{github.com/empymod/article-fdesign}.
The examples can be used as templates to design new filters.

@@

The main features of \texttt{fdesign} and some differences between it and the
methods it is based upon are:
\begin{itemize}
  \item The algorithm computes many different filters for various spacing and
    shift values (brute-force), with an optional minimization routine for
    the best outcome of the brute-force step.
  \item \cite{GP.07.Kong} and \cite{GEO.12.Key} optimize a filter for the $J_0$
    filter, and then use the obtained shift and spacing value to calculate the
    $J_1$ values. The presented algorithm can optimize either $J_0$, $J_1$, or
    both in the optimization process. Five transform-pairs for each $J_0$ and
    $J_1$, and three transform pairs for each sine and cosine of often used
    functions are included in the routine, but any other transform pair can be
    provided as input.
  \item More complex or more specific numerical transform  pairs  can be
    provided instead of theoretical transform pairs by using the modeler
    \texttt{empymod}.
  \item \cite{GP.07.Kong} defines a \emph{good} filter as one that recovers
    \emph{weak} diffusive EM fields. In  \texttt{fdesign}, you can define a
    relative error level that defines up to what error an obtained result is
    good or not. The optimization seeks to find the smallest amplitude field
    that is accurately modeled (i.e., minimizing the amplitude), but it has
    additionally another mode, where it will maximize the output abscissae $r$
    of the transform pair instead of minimizing the amplitude. The two modes
    yield the same result in the general, simple case of fast decaying
    transform pairs. However, for some complex models and specifically for very
    high frequencies, maximizing $r$ will yield much better and more consistent
    results.
  \item The algorithm can solve under-, equal-, and over-determined linear
    systems.
  \item The real or the imaginary part can be used for the inversion of complex
    transform pairs.
\end{itemize}

@@

\subsection{Minimization criteria}
Figure~\ref{fig:InvCrit} shows the differences between the different
minimization approaches; in (a) for a conventional, fast decaying transform
pair; and in (b) for a more complex, high frequency layered model. The circles
show the minimum amplitude used in previous approaches. This criterion is not
ideal, as it is subject to some random fluctuations and depends on the choice
of $r$. The squares show the minimum amplitude given a certain acceptable
error, and the diamonds show the maximum $r$ given a certain acceptable error
(the inversion is a minimization process, it therefore minimizes $1/r$, not
$r$). In simple cases the minimum amplitude given an acceptable error and the
maximum $r$ given an acceptable error will yield the same result. However, in
complex cases the maximum $r$ is more consistent and therefore a much better
criterion.
%
\plot*{InvCrit}{width=.9\textwidth}{(a) A regular RHS curve of a transform pair
  with a purely decaying function. Minimum amplitude or maximum $r$ yield the
  same result in this scenario. Using a relative error criteria is more stable
  than just the absolute minimum amplitude. (b) A 1D model for $f=200\,$MHz.
  Here, the maximum $r$ provides a better criteria for the inversion.}
%

@@

\subsection{Design}
Figure~\ref{fig:201} shows the solution spaces of four consecutive inversion
runs, where each run is a focus on a subsection of the previous run, indicated
by the red square.
%
\plot*{201}{width=\textwidth}{%
  Solution spaces of four consecutive inversion runs for a 201\,pt filter. Each
  consecutive run zooms into a portion of the previous solution space,
  indicated by the red square. The more in detail we obtain the solution space,
  the more random appears to be the distribution.}%
%
The filter length in this case is $N=201$, and the theoretical transform pairs
are as given, e.g., in \cite{USGS.75.Anderson}:
%
\begin{align}
  \int^\infty_0\,l \exp\left(-al^2\right) J_0(rl)\,dl &=
  \frac{\exp\left(\frac{-r^2}{4a}\right)}{2a}\ , \\
  \int^\infty_0\,l^2 \exp\left(-al^2\right) J_1(rl)\,dl &=
  \frac{r}{4a^2} \exp\left(-\frac{r^2}{4a}\right)\ ,
  \label{eq:j01}
\end{align}
%
where $a$ was set to 5. In the algorithm you can provide one pair for the
inversion, and a different pair to get the minimum amplitude or the maximum
$r$. In this case the same transform pair was used for both the inversion and
the check of quality. The acceptable relative error was set to 1\,\%. The RHS
evaluation parameter for the inversion was set so  that an over-determined
system was evaluated with $M = 2N$ equations, where $r_\mr{I}$ was
logarithmically spaced from $\log_{10}(1/\max(b)) - 1$ to $\log_{10}(1/\min(b))
+ 1$, where $b$ are the filter abscissae as given in equation~\ref{eq:base}.
This is accomplished in \texttt{fdesign} by setting parameter $r_\mr{def} = (1,
1, 2)$.

From the low resolution overview runs (a) and (b) it looks like this is a
standard, straight-forward minimization problem. However, from the more
detailed results in (c) and (d) it becomes obvious that it is a minimization
problem that has to be solved stochastically, as there are solutions with 2
orders of magnitude difference apparently randomly next to each other.

@@

Designing filters is to a large extent trial and error. All input variables
influence the outcome, and often  a good filter is found by sheer luck of
choosing the right starting parameters. Each of the input parameters has an
effect on the outcome. It also depends heavily on the transform pairs
$f_\mr{I;C}$; functions that decay rapidly are generally better, as noted by
earlier authors \citep[e.g.][]{USGS.75.Anderson}. It also depends on the filter
length, and obviously heavily on the spacing and shift values. Another
important point is how you define the right-hand evaluation points of the
inversion ($r_\mr{def}$). Evaluating corresponding transform pairs separately
or jointly also leads to different filter coefficients ($J_0$, $J_1$, or $J_0$
\& $J_1$, or equally sine, cosine, or sine \& cosine); also if the real or the
imaginary part is used when complex transform pairs or \texttt{empymod} is
used; and if it is inverted for the minimum amplitude or for the maximum $r$.

@@

It is very important to note again that other scenarios might yield very
different error plots. Although the new 201\,pt filter proves to be very
accurate for these three models, it might not be the best filter in other
cases. To verify the applicability of the filter we run two different tests
over a wide range of scenarios. Figure~\ref{fig:error} shows the relative
errors of DLF using the filter \emph{Wer201} compared to QWE for a deep water
model, a shallow water model, and a land model. The models consist of air
(upper halfspace), a water layer in the first two examples, and a subsurface
halfspace. Source depth is 1990\,m, 10\,m, and 0.5\,m in the deep, shallow, and
land case, respectively. Receiver depth is 2000\,m, 100\,m, and 0.8\,m. The
subsurface halfspace resistivity varies from 1 to 10, 100, and
1000\,$\Omega\,$m from the left to the right column. The error plots cover
offsets from 50\,m to 20\,km on the x-axis, and 0.01\,Hz to 10\,Hz on the
y-axis. It can be seen in the result that the filter is sufficiently precise
for all practical CSEM cases, hence has generally an error far below 1\,\%
($10^{-2}$) in all cases. The exception are large offsets with relatively high
frequencies, e.g., 10\,km offset with 8\,Hz. In the shallow water case this
zone has a very high error for subsurface resistivities of 1\,$\Omega\,$m and
10\,$\Omega\,$m (yellow zones). However, this is due to the fact that the
amplitudes of the signal for these offsets and frequencies are in the order of
$10^{-25}\,$V/m.
%
\plot*{error}{width=\textwidth}{Relative error plots comparing the DLF
  (\emph{Wer201}) with QWE for the three cases deep water (top row), shallow
  water (middle row), and land (bottom row). The subsurface halfspace
  resistivity varies from 1 to 1000\,$\Omega\,$m from the left column to the
  right column; offsets vary from 50\,m to 20\,km, and frequencies vary from
  0.01\,Hz to 10\,Hz. The DLF provides sufficiently accurate results for all
  practical CSEM applications. The exception is in the shallow water case for
  $\rho = 1\,\Omega\,$m and $10\,\Omega\,$m for large offsets and high
  frequencies (yellow zones). However, the amplitudes in these regions are in
  the order of $10^{-25}\,$V/m, and therefore far below the noise level.}
%

>>>>

In the following sections we give a brief overview of the methodology, the
theory, and the code, and then show examples of its usage, as well as
presenting new filters for both CSEM and GPR data. The current article falls
therefore into all three mentioned categories by the new application of DLF to
wave phenomena, \new{which means that the DLF can probably be used to model
seismic wavefields as well,} and by providing new filters with the accompanying
code. The algorithm and many more examples of its usage can be found on
\href{https://github.com/empymod/article-fdesign}{github.com/empymod/article-fdesign}.
The examples can be used as templates to design new filters.

@@

The main features of \texttt{fdesign} and some differences between it and the
methods it is based upon are:
\begin{itemize}
  \item The algorithm computes many different filters for various spacing and
    shift values (brute-force), with an optional minimization routine for
    the best outcome of the brute-force step.
  \item \cite{GP.07.Kong} and \cite{GEO.12.Key} optimize a filter for the $J_0$
    filter, and then use the obtained shift and spacing value to calculate the
    $J_1$ values. The presented algorithm can optimize either $J_0$, $J_1$, or
    both in the optimization process. Five transform-pairs for each $J_0$ and
    $J_1$, and three transform pairs for each sine and cosine of often used
    functions are included in the routine, but any other transform pair can be
    provided as input.
  \item More complex or more specific numerical transform  pairs  can be
    provided instead of theoretical transform pairs by using the modeler
    \texttt{empymod}.
  \item \cite{GP.07.Kong} defines a \emph{good} filter as one that recovers
    \emph{weak} diffusive EM fields. In  \texttt{fdesign}, you can define a
    relative error level that defines up to what error an obtained result is
    good or not. The optimization seeks to find the smallest amplitude field
    that is accurately modeled (i.e., minimizing the amplitude), but it has
    additionally another mode, where it will maximize the output abscissae $r$
    of the transform pair instead of minimizing the amplitude. The two modes
    yield the same result in the general, simple case of fast decaying
    transform pairs. However, for some complex models and specifically for very
    high frequencies, maximizing $r$ will yield much better and more consistent
    results.
  \item The algorithm can solve \old{under-,} equal-\old{,} and over-determined
    linear systems.
  \item The real or the imaginary part can be used for the inversion of complex
    transform pairs.
\end{itemize}

@@

\new{To solve step 2(c) we implemented the QR-decomposition method in
  \texttt{fdesign}.}

\subsection{Minimization criteria}
Figure~\ref{fig:InvCrit} shows the differences between the different
minimization approaches; in (a) for a conventional, fast decaying transform
pair \new{(using the $J_0$ kernel of a fullspace with vertical distance between
source and receiver of 50\,m, frequency $f=1\,$Hz, and resistivity
$\rho=1\,\Omega\,$m)}; and in (b) for a more complex, high frequency layered
model \new{(using the $J_0$ kernel for $f=200\,$MHz for the GPR model described
in the section \emph{Ground-penetrating radar}, with receiver located at 1\,m
depth)}. The circles show the minimum amplitude used in previous approaches.
This criterion is not ideal, as it is subject to some random fluctuations and
depends on the choice of $r$. The squares show the minimum amplitude given a
certain acceptable error, and the diamonds show the maximum $r$ given a certain
acceptable error (the inversion is a minimization process, it therefore
minimizes $1/r$, not $r$). In simple cases the minimum amplitude given an
acceptable error and the maximum $r$ given an acceptable error will yield the
same result. However, in complex cases the maximum $r$ is more consistent and
therefore a much better criterion.
%
\plot*{InvCrit}{width=.9\textwidth}{(a) A regular RHS curve of a transform pair
  with a purely decaying function. Minimum amplitude or maximum $r$ yield the
  same result in this scenario. Using a relative error criteria is more stable
  than just the absolute minimum amplitude. (b) A 1D model for $f=200\,$MHz.
  Here, the maximum $r$ provides a better criteria for the inversion.}
%

@@

\subsection{Design}
Figure~\ref{fig:201} shows the solution spaces of \old{four}\new{eight}
consecutive inversion runs\old{, where each run is a focus on a subsection of
  the previous run, indicated by the red square. The filter length in this case
is $N=201$, and}\new{where} the theoretical transform pairs
\old{are} as given, e.g., in \cite{USGS.75.Anderson}\old{:}\new{,}
%
\begin{align}
  \int^\infty_0\,l \exp\left(-al^2\right) J_0(rl)\,dl &=
  \frac{\exp\left(\frac{-r^2}{4a}\right)}{2a}\ , \\
  \int^\infty_0\,l^2 \exp\left(-al^2\right) J_1(rl)\,dl &=
  \frac{r}{4a^2} \exp\left(-\frac{r^2}{4a}\right)\ ,
  \label{eq:j01}
\end{align}
%
%
\plot*{201}{width=\textwidth}{%
  Solution spaces of four consecutive inversion runs for a 201\,pt filter. Each
  consecutive run zooms into a portion of the previous solution space,
  indicated by the red square. The more in detail we obtain the solution space,
  the more random appears to be the distribution.}%
%
where \new{used; } $a$ was set to 5. In the algorithm you can provide one pair
for the inversion, and a different pair to get the minimum amplitude or the
maximum $r$. In this case the same transform pair was used for both the
inversion and the check of quality. The acceptable relative error was set to
1\,\%. The RHS evaluation parameter for the inversion was set so  that an
over-determined system was evaluated with $M = 2N$ equations, where $r_\mr{I}$
was logarithmically spaced from $\log_{10}(1/\max(b)) - 1$ to
$\log_{10}(1/\min(b)) + 1$, where $b$ are the filter abscissae as given in
equation~\ref{eq:base}. This is accomplished in \texttt{fdesign} by setting
parameter $r_\mr{def} = (1, 1, 2)$.

\new{The top row shows in subplots (a) to (d) four different inversion runs
for $N=51$, 101, 201, and 401 for a wide range of spacing and shift values.
From this one has to decide for a filter length. Longer filters tend to give
more precise results, but their transform is computationally more expensive as
more values have to be evaluated. The length of the filter does not has to be
an odd number, but many published filters do have an odd number of
coefficients. The integral runs from $-\infty$ to $+\infty$ and hence an odd
point filter comes naturally by truncating the integral to a finite sum from
$-(N-1)/2$ to $(N-1)/2$. However, this is only true if the shift value is
$\delta = 0$. If your shift value is not zero then there is no advantage in
having an odd point filter.}

\new{The bottom row shows four consecutive inversion runs for $N=201$, where
each run is a focus on a subsection of the previous run, indicated by the red
square.} From the low resolution overview runs (e) and (f) it looks like this
is a standard, straight-forward minimization problem. However, from the more
detailed results in (g) and (h) it becomes obvious that it is a minimization
problem that has to be solved stochastically, as there are solutions with 2
orders of magnitude difference apparently randomly next to each other.

@@

Designing filters is to a large extent trial and error. All input variables
influence the outcome, and often a good filter is found by sheer luck of
choosing the right starting parameters\new{, as can be seen in
Figure~\ref{fig:201b} (h). However, once you are in this zone every filter is a
very good filter}. Each of the input parameters has an effect on the outcome.
It also depends heavily on the transform pairs $f_\mr{I;C}$; functions that
decay rapidly are generally better, as noted by earlier authors
\citep[e.g.][]{USGS.75.Anderson}. It also depends on the filter length, and
obviously heavily on the spacing and shift values. Another important point is
how you define the right-hand evaluation points of the inversion
($r_\mr{def}$). Evaluating corresponding transform pairs separately or jointly
also leads to different filter coefficients ($J_0$, $J_1$, or $J_0$ \& $J_1$,
or equally sine, cosine, or sine \& cosine); also if the real or the imaginary
part is used when complex transform pairs or \texttt{empymod} is used; and if
it is inverted for the minimum amplitude or for the maximum $r$.

@@

It is very important to note again that other scenarios might yield very
different error plots. Although the new 201\,pt filter proves to be very
accurate for these three models, it might not be the best filter in other
cases. To verify the applicability of the filter we run two different tests
over a wide range of scenarios. Figure~\ref{fig:error} shows the relative
errors of DLF using the filter \emph{Wer201} compared to QWE for a deep water
model, a shallow water model, and a land model. The models consist of air
(upper halfspace), a water layer in the first two examples, and a subsurface
halfspace. Source depth is 1990\,m, 10\,m, and 0.5\,m in the deep, shallow, and
land case, respectively. Receiver depth is 2000\,m, 100\,m, and 0.8\,m. The
subsurface halfspace resistivity varies from 1 to 10, 100, and
1000\,$\Omega\,$m from the left to the right column. The error plots cover
offsets from 50\,m to 20\,km on the x-axis, and 0.01\,Hz to 10\,Hz on the
y-axis. It can be seen in the result that the filter is sufficiently precise
for all practical CSEM cases, hence has generally an error far below 1\,\%
($10^{-2}$) in all cases. The exception are large offsets with relatively high
frequencies, e.g., 10\,km offset with 8\,Hz. In the shallow water case this
zone has a very high error for subsurface resistivities of 1\,$\Omega\,$m and
10\,$\Omega\,$m (yellow zones). However, this is due to the fact that the
amplitudes of the signal for these offsets and frequencies are in the order of
$10^{-25}\,$V/m.
%
\plot*{error}{width=\textwidth}{Relative error plots comparing the DLF
  (\emph{Wer201}) with QWE for the three cases deep water (top row), shallow
  water (middle row), and land (bottom row). The subsurface halfspace
  resistivity varies from 1 to 1000\,$\Omega\,$m from the left column to the
  right column; offsets vary from 50\,m to 20\,km, and frequencies vary from
  0.01\,Hz to 10\,Hz. The DLF provides sufficiently accurate results for all
  practical CSEM applications. The exception is in the shallow water case for
  $\rho = 1\,\Omega\,$m and $10\,\Omega\,$m for large offsets and high
  frequencies (yellow zones). However, the amplitudes in these regions are in
  the order of $10^{-25}\,$V/m, and therefore \old{far}\new{several orders of
  magnitude} below the noise level\new{s of current EM instrument systems}.}
%
